---
title: "Estimation on simulated data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{estimation_on_simulated_data_simple}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
rmarkdown.html_vignette.check_title = FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Recap

dvsb estimates **d**isease **v**ariability in **s**ubpopulations from **b**iomarkers, e.g. estimating the variability between subpopulations in seroprevalence (the proportion of people who are seropositive) with antibody levels estimated using an assay measuring some proxy such as ELISA Optical Density values.
We call the biomarker x (e.g. antibody level) and the proxy measurement y (e.g. OD values).

## Setting up

```{r setup, message=FALSE}
library(dvsb)
library(tidyr)
library(dplyr)
library(ggplot2)
theme_set(theme_classic())
```

## Simulate data

We simulate data, giving us a list with seven things in it.
(If we were using real data instead of simulated data, we would simply need to create the `df_sam` and `df_cal` dataframes ourselves instead of running `simulate_data()` below.
The documentation for `wrangle_data()` explains what columns these dataframes require, and how sets of covariates for regression models are specified using the arguments `x_mix_pred_vars_names`, `p_pos_binary_pred_vars`, `f_pred_vars_names`.)

```{r simulate, message=FALSE}
data <- simulate_data(seed = 1234,
                      num_sam_per_plate = 100)
df_sam <- data$df_sam
df_cal <- data$df_cal
df_plate <- data$df_plate
param_true_values_list <- data$params
f_pred_vars_names <- data$f_pred_vars_names
x_mix_pred_vars_names <- data$x_mix_pred_vars_names
p_pos_binary_pred_vars <- data$p_pos_binary_pred_vars
```

Plot the simulated calibration data (one facet per plate):

```{r plot_data_cal, fig.height=5}
ggplot(df_cal) +
  geom_point(aes(x = x, y = y)) +
  facet_wrap(~plate, ncol = 3) 
```

Plot the simulated sample data, histogramming the x and y values, one facet for each category of the `letter` covariate (which differ in their x distributions when using default arguments for `simulate_data()`):

```{r plot_data_sam_x, fig.height=5}
ggplot(df_sam) +
  geom_histogram(aes(x = x), bins = 20) +
  facet_wrap(~letter) +
  scale_x_log10()
```

```{r plot_data_sam_y, fig.height=5}
# With default values, y is rarely negative; restrict to positive for visibility
ggplot(df_sam %>% filter(y > 0)) +
  geom_histogram(aes(x = y), bins = 20) +
  facet_wrap(~letter) +
  scale_x_log10()
```

## Run Stan

Get the default prior distributions.

``` {r get_priors}
priors_list <- get_priors()
```

Wrangle the data and priors into the form needed for stan.

``` {r wrangle}
data_wrangled <- wrangle_data(
  df_sam = df_sam, 
  df_cal = df_cal, 
  df_priors_scalars = priors_list$df_priors_scalars,
  df_priors_vectors = priors_list$df_priors_vectors, 
  rho_prior_eta = priors_list$rho_prior_eta, 
  x_mix_pred_vars_names = x_mix_pred_vars_names, 
  p_pos_binary_pred_vars = p_pos_binary_pred_vars,
  f_pred_vars_names = f_pred_vars_names
  )
```

Update our dataframes for samples, calibrators and plates with the extra columns we just added (which introduce integer indexing); this can be helpful later if we want to combine results and data.

``` {r}
df_sam <- data_wrangled$df_sam
df_cal <- data_wrangled$df_cal
df_plate <- dplyr::left_join(df_plate, data_wrangled$df_plate, by = "plate") 
```

We run Stan once to sample from the posterior, and again to sample from the prior.
(We wrap inside `capture.output( )` to suppress Stan's insistent printing of information.)

```{r run_stan, error = FALSE, warning = FALSE, message = FALSE, results = "hide"}
capture.output({
df_posterior <- run_stan(
  data_wrangled = data_wrangled,
  distribution = "posterior")

df_prior <- run_stan(
  data_wrangled = data_wrangled,
  distribution = "prior")
})
```

## Analyse Stan output

`df_posterior` and `df_prior` are the dataframes (in data.table format) of samples, from the posterior and prior respectively, as returned by Stan.

We wrangle the true parameter values - flattening the list of values into a vector and renaming - for plotting convenience:

``` {r wrangle_true_params}
param_true_values_vector <- wrangle_true_params(
  param_true_values_list = param_true_values_list,
  data_descriptors = data_wrangled$data_descriptors)
```

For each of the population-level parameters we plot the posterior, prior and true value, demonstrating that the posterior moves away from the prior toward the true value, as it should for a correctly implemented model with informative data.
One should always test this.
We do this using `plot_posterior()` from our package `mastiff`, just for nine parameters for better visibility within Rmarkdown:

```{r plot_posterior_prior, warning=FALSE, out.width="95%", fig.height=10}
mastiff::plot_posterior(
    posterior_samples = df_posterior,
    prior_samples     = df_prior,
    params_desired = names(df_prior)[1:9],
    true_param_values = param_true_values_vector,
    skip_stanfit_to_dt = TRUE)
```

`df_posterior` has columns named `xlog_sam[1]`, `xlog_sam[2]` etc. which contain the posterior samples for x (on a log scale) for the sample assigned the integer index 1, and 2, etc.
Calculate some quantiles of the posterior for each of the samples, and compare these to the true values which we had in `df_sam`.
The smallest values are slightly overestimated and the largest values are slightly underestimated: this is the phenomenon of shrinkage toward the mean, due to these parameters being random effects.

```{r plot_x_vs_true, message = FALSE, warning = FALSE}
quantiles <- c(0.025, 0.5, 0.975)
df_sam_x <- df_posterior %>%
  select(starts_with("xlog_sam[")) %>%
  pivot_longer(everything(), names_to = "param") %>%
  extract(param,
          into = "id_sam_int", 
          regex = "xlog_sam\\[([0-9]+)\\]") %>%
  mutate(id_sam_int = as.integer(id_sam_int)) %>%
  group_by(id_sam_int) %>%
  reframe(value = quantile(value, probs = quantiles),
          quantile = quantiles) %>%
  pivot_wider(names_from = quantile, names_prefix = "x_quant_") %>%
  left_join(df_sam %>%
              select(id_sam_int, xlog) %>%
              distinct(),
            by = "id_sam_int")
ggplot(df_sam_x) +
  geom_errorbar(aes(xlog, ymin = x_quant_0.025, ymax = x_quant_0.975)) +
  geom_point(aes(xlog, x_quant_0.5)) +
  geom_abline() +
  labs(x = "True log x",
       y = "Estimated log x")
```

`df_posterior` has columns named `y_sam_sim_unconditional[1]`, `y_sam_sim_unconditional[2]` etc. which contain posterior predicted y values for new set of sampled individuals with the same distribution of covariates (i.e. partitioned into the same subpopulations, in this example based on a single covariate `letter`) as in the dataset analysed.
The integer in square brackets indexes the sample replicate as in `df_sam$which_sam_rep`; we merge `df_sam` into `df_posterior` to associated each predicted `y` value with the appropriate subpopulation.
We plot the empirical subpopulation-level distributions of y values, and the posterior predictive distribution of subpopulation-level distributions of y values.
We use a log scale for y even though it can be negative, because with our default parameters it's only rarely negative and a log scale is clearer.
We make both a spline density plot and also a histogram (splines are easier to see but the algorithm may make mistakes on fitting a smoothed distribution).

```{r plot_posterior_retrodictive, message = FALSE, warning = FALSE}
df_y_sim <- df_posterior %>%
  select(starts_with("y_sam_sim_unconditional[")) %>%
  mutate(sample = row_number()) %>%
  pivot_longer(-c("sample"), names_to = "param") %>%
  tidyr::extract(param, 
               into = "which_sam_rep", 
               regex = "y_sam_sim_unconditional\\[([0-9]+)\\]") %>%
  mutate(which_sam_rep = as.integer(which_sam_rep)) %>%
  left_join(df_sam, by = "which_sam_rep") %>%
  mutate(value = log10(value))
ggplot() +
  geom_step(data = df_y_sim,
            aes(x = value,
                y = after_stat(density),
                group = sample),
            stat="bin",
            color="black",
            bins = 60,
            alpha = 0.1) +
  geom_step(data = df_sam,
            aes(x = log10(y), # y,
                y = after_stat(density)),
            stat="bin",
            bins = 60,
            color="blue") +
  coord_cartesian(expand = F) +
  facet_wrap(~letter) +
  labs(x = "log10(OD value)",
       y = "probability density")
ggplot() +
  geom_density(data = df_y_sim,
               aes(x = value,
                   group = sample),
               color = "black") +
  geom_density(data = df_sam,
            aes(x = log10(y)),
            color="blue") +
  coord_cartesian(expand = F) +
  facet_wrap(~letter) +
  labs(x = "log10(OD value)",
       y = "probability density")
```

`df_posterior` has columns named `p_sam_is_pos[1]`, `p_sam_is_pos[2]` etc. which contain the probability that sample 1, 2 etc. is positive given its observed y values. 
Calculate the posterior mean for each of these quantities, and then plot the distribution over individuals of these values, stratified by true serostatus.
This gives a sense of the model's accuracy and confidence in estimating serostatus. 

```{r, warning = FALSE}
df_prob_pos <- df_posterior %>%
  select(starts_with("p_sam_is_pos["))
df_prob_pos <- df_prob_pos[, lapply(.SD, mean)]
df_prob_pos <- df_prob_pos %>%
  pivot_longer(everything(), names_to = "param", values_to = "p_sam_is_pos") %>%
  tidyr::extract(param, 
                 into = "id_sam_int", 
                 regex = "p_sam_is_pos\\[([0-9]+)\\]") %>%
  mutate(id_sam_int = as.integer(id_sam_int)) %>%
  inner_join(df_sam %>% summarise(.by = id_sam_int, pos = unique(pos)), 
             by = "id_sam_int") %>%
    mutate(status = if_else(pos, "pos", "neg")) 

ggplot(df_prob_pos) +
  geom_histogram(aes(x = p_sam_is_pos, y = after_stat(density), fill = status),
                 position = "identity", alpha = 0.6, bins = 30) +
  labs(y = "Number of samples",
       x = "Probability sample is positive") +
  coord_cartesian(expand = FALSE) +
  scale_x_continuous(limits = c(NA, NA))
```

